# Geodata products {#Ch05}

Some raw data need extensive processing before EGVs can be created, many EGVs 
depend on processing raw geodata into geodata products before EGVs can be create, 
and in some cases, EGV itself could be created from ras geodata, but it has 
to be spatially restricted to certain locations. This chapter describes these geodata 
products and procedures involved in creating them.

## Terrain products {#Ch05.01}

In order to develop part of the relief-related EGV, such as the topographic 
moisture index (TWI) and non-drainage depressions, it is necessary to address 
water flow in the environment. This is a multi-step procedure that is logical 
and reliable in mountainous areas and environments with little hydrological 
impact. However, in the context of Latvia, this is challenging. These challenges 
can be addressed in various ways. For example, if reliable (accurate) information 
on the exact locations of rivers and ditches were available, it could be 
incorporated into the terrain. Unfortunately, there is no sufficiently 
accurate information available. Therefore, information about transport 
structures from the [Melioration Cadastre Information System database](#Ch04.03), bridges 
from the [topographic map](#Ch04.04) and transport structures and bridges 
from [LVM Open Data](#Ch04.06) was used to address the challenges - 
information about the minimum height above sea level was incorporated into the 
DEM to be used further in a 30 m buffer around these locations.

```{r, eval=FALSE}
# Libs

# arī LVM

```

This DEM was then used for geoprocessing to find terrain depressions and 
determine the topographic wetness index (TWI). First, terrain depressions were 
calculated, assuming that all runoff that should be known is known and taken 
into account (adjusted DEM values - previous code area). Then, the topographic 
wetness index was prepared and the search for non-runoff depressions was repeated:

1. to calculate the topographic wetness index, the terrain depressions without 
runoff were reviewed, allowing up to ten cell breaks in places of lower 
resistance, the rest were filled in;

2. precise drainage depressions and their depth layers were prepared after 
incorporating flow breaks;

3. for additional security, the result of the first step was repeated to 
search for and fill in relief depressions [@WangLiu2006];

4. the result of the third step was used to determine the specific catchment 
area using d-infinity flow division;

5. by combining the specific catchment area layer with the slope layer, 
the topographic wetness index was calculated. A graphical evaluation revealed 
individual extreme values, which were limited to **20**.


```{r, eval=FALSE}
# Libs
```

Since the initial DEM input was created by filling in water bodies using 
interpolation methods, the water bodies show a pronounced terrain, which needs 
to be removed. This is done by inserting average values into these polygons.


```{r, eval=FALSE}
# Libs
```



## Soil texture product {#Ch05.02}

In this section one united layer describing categorised soil texture (sand=1, 
silt=2, clay=3, organic=4) is created from multiple preprocessed soil texture 
data sources. Creation of soil texture product consisted of multiple overlay steps. 
These steps are illustrated together with processed geodata used:

1. the basis soil texture source was [Soil texture from the European Soil Database](#Ch04.07.02), 
this layer had to be reclassified to match other layers as it was not performed 
during preprocessing;

2. the layer from the first step was overlaid by [Latvian Quarternary geology data](#Ch04.07.04) 
written as numeric starting with 1;

3. the layer from the second step was overlaid by [20th century topsoil in Latvian farmland](#Ch04.07.03) 
written as numeric starting with 1;

4. the layer from [Organic soils as modelled by Silava](#Ch04.07.05) (presence-only) 
was overlaid by [Organic soils as modelled by University of Latvia](#Ch04.07.06) 
(presence-absence). After the overlay, it was classified as presence-only;

5. the layer from the third step was overlaid by the layer from the fourth and 
saved for EGV creation.


```{r,eval=FALSE}
# libs ----
if(!require(terra)) {install.packages("terra"); require(terra)}

# step 1
step1=rast("./RasterGrids_10m/2024/SoilTXT_ESDAC.tif")
step1x=ifel(step1==1,1,
            ifel(step1==2,2,
                 ifel(step1==3,2,
                      ifel(step1==4,3,
                           ifel(step1==8,4,NA)))))
plot(step1x)
step1xy=as.numeric(step1x)
plot(step1xy)


# step 2
step2a=rast("./RasterGrids_10m/2024/SoilTXT_QuarternaryLV.tif")
step2a=as.numeric(step2a)+1
plot(step2a)

step2=cover(step2a,step1x)
plot(step2)

# step 3
step3a=rast("./RasterGrids_10m/2024/SoilTXT_topSoilLV.tif")
step3a=as.numeric(step3a)+1
plot(step3a)

step3=cover(step3a,step2)
plot(step3)

# step 4
step4a=rast("./RasterGrids_10m/2024/SoilTXT_OrganicLU.tif")
step4b=rast("./RasterGrids_10m/2024/SoilTXT_OrganicSilava.tif")

step4c=cover(step4a,step4b)

step4=ifel(step4c==1,4,NA)
plot(step4)

# step 5

step5=cover(step4,step3)
plot(step5)

writeRaster(step5,
           "./RasterGrids_10m/2024/SoilTXT_combined.tif",
           overwrite=TRUE)


```



## Landscape classification {#Ch05.03}

In this exercise, "landscape" refers to the representation of different types 
of land cover and land use classes, where the order in which these classes are 
drawn is important, because spatial data from different sources often have 
mismatched boundaries, which requires addressing both their overlap (1) and 
filling in gaps where there is no database information (2), and the choice of 
how to emphasize objects with certain processing, such as buffering, because 
some elements that are important for characterizing the environment (especially 
edge effects) may be so small or so poorly positioned that they disappear during 
the rasterization process (3). The general landscape layer also serves as a 
mask for the preparation of further environmental descriptions. This section 
describes the development of a general (simple) landscape and, in the following 
document, its enrichment with more specific environmental eco-geographical 
variables. The general landscape is stored in the file `Ainava_vienk_mask.tif`, 
in which the classes and the procedure for their creation are described in the 
following list:

- class `100` - **roads**: roads from various sources, **filled in sequence** - 
dominates classes with higher values so that relatively small objects are not 
lost and information about edges is provided. The following have been 
combined to create this class:

    - layers `RoadA_COMB` and `RoadL_COMB` (except smallest size groups) from
    [topographic map](#Ch04.04), buffered by 10 m before rasterization;
    
    - [LVM open data](#Ch04.06) layers `LVM_MEZA_AUTOCELI`, `LVM_ATTISTAMIE_AUTOCELI`, 
    `LVM_APGRIESANAS_LAUKUMI`, `LVM_IZMAINISANAS_VIETAS` and `LVM_NOBRAUKTUVES` 
    buffered by 10 m;
    
    - information from the State Forest Register on natural roads has not 
    been used, as these do not usually form a continuous break in the canopy. 
    Information on roads from this register is also available in other 
    resources and has not been duplicated.

The command lines below create a layer with landscape class `100`, which is 
saved in the file `SimpleLandscape_class100_celi.tif` for further processing.


```{r, eval=FALSE}
# Libs ----
if(!require(tidyverse)) {install.packages("tidyverse"); require(tidyverse)}
if(!require(sf)) {install.packages("sf"); require(sf)}
if(!require(arrow)) {install.packages("arrow"); require(arrow)}
if(!require(sfarrow)) {install.packages("sfarrow"); require(sfarrow)}
if(!require(terra)) {install.packages("terra"); require(terra)}
if(!require(raster)) {install.packages("raster"); require(raster)}
if(!require(fasterize)) {install.packages("fasterize"); require(fasterize)}
if(!require(gdalUtilities)){install.packages("gdalUtilities");require(gdalUtilities)}
if(!require(readxl)) {install.packages("readxl"); require(readxl)}

# templates ----
template_t=rast("./Templates/TemplateRasters/LV10m_10km.tif")
template_r=raster(template_t)


# class 100 ----

#poly
celi_topo=st_read_parquet("./Geodata/2024/TopographicMap/RoadA_COMB.parquet")
celi_topo=celi_topo %>% 
  mutate(yes=100) %>% 
  dplyr::select(yes)
ctb=st_buffer(celi_topo,dist=10)
r_celi_topo=fasterize(ctb,template_r,field="yes")

# pts
nobrauktuves=st_read("./Geodata/2024/LVM_OpenData/LVM_NOBRAUKTUVES/LVM_NOBRAUKTUVES_Shape.shp")
nobrauktuves=nobrauktuves %>% 
  mutate(yes=100) %>% 
  dplyr::select(yes)
izmainisanas=st_read("./Geodata/2024/LVM_OpenData/LVM_IZMAINISANAS_VIETAS/LVM_IZMAINISANAS_VIETAS_Shape.shp")
izmainisanas=izmainisanas %>% 
  mutate(yes=100) %>% 
  dplyr::select(yes)
apgriesanas=st_read("./Geodata/2024/LVM_OpenData/LVM_APGRIESANAS_LAUKUMI/LVM_APGRIESANAS_LAUKUMI_Shape.shp")
apgriesanas=apgriesanas %>% 
  mutate(yes=100) %>% 
  dplyr::select(yes)
cp=rbind(nobrauktuves,izmainisanas,apgriesanas)
cpb=st_buffer(cp,dist=10)
r_celi_pts=fasterize(cpb,template_r,field="yes")


# lines
meza_autoceli=st_read("./Geodata/2024/LVM_OpenData/LVM_MEZA_AUTOCELI/LVM_MEZA_AUTOCELI_Shape.shp")
meza_autoceli=meza_autoceli %>% 
  mutate(yes=100) %>% 
  dplyr::select(yes)
attistamie=st_read("./Geodata/2024/LVM_OpenData/LVM_ATTISTAMIE_AUTOCELI/LVM_ATTISTAMIE_AUTOCELI_Shape.shp")
attistamie=attistamie %>% 
  mutate(yes=100) %>% 
  dplyr::select(yes)
topo_lines=st_read_parquet("./Geodata/2024/TopographicMap/RoadL_COMB.parquet")
topo_lines=topo_lines %>% 
  mutate(yes=100) %>% 
  dplyr::select(yes)
cl=bind_rows(meza_autoceli,attistamie,topo_lines)
cl=cl %>% 
  dplyr::select(yes)
clb=st_buffer(cl,dist=10)
r_celi_lines=fasterize(clb,template_r,field="yes")

# cleaning
rm(apgriesanas)
rm(attistamie)
rm(celi_topo)
rm(topo_lines)
rm(ctb)
rm(cl)
rm(clb)
rm(cp)
rm(cpb)
rm(izmainisanas)
rm(meza_autoceli)
rm(nobrauktuves)

# to terra
t_celi_topo=rast(r_celi_topo)
t_celi_pts=rast(r_celi_pts)
t_celi_lines=rast(r_celi_lines)

# cleaning
rm(r_celi_lines)
rm(r_celi_pts)
rm(r_celi_topo)

# union
plot(t_celi_topo)

road_union1=cover(t_celi_topo,t_celi_pts)
road_union2=cover(road_union1,t_celi_lines,
                  filename="./RasterGrids_10m/2024/SimpleLandscape_class100_celi.tif",
                  overwrite=TRUE)

# cleaning
rm(t_celi_topo)
rm(t_celi_pts)
rm(t_celi_lines)
rm(road_union1)
rm(road_union2)

```

- class `200` - **waters**: water bodies from various sources, filled in 
sequence (but see "merging and filling" step of this section) - dominates classes 
with higher values so that relatively small objects are not lost and information 
about the edges is provided. The following were combined to create this class:

– [topographic map](#Ch04.04) layers `HidroA_COMB` and `HidroL_COMB` (buffered by 5 m);

– [MKIS](#Ch04.03) layer `Gravji`, buffered by 3 m;

– [LVM open data](#Ch04.06) layers `LVM_GRAVJI`, buffered by 5 m.

– Information about ditches from the State Forest Register has not been used, 
as it is also available in other resources, or is of so small structures that it does 
not cause a continuous break in the tree canopy.

The command lines below create a layer with landscape class `200`, which is 
saved in the file `SimpleLandscape_class200_udens_premask.tif` for further processing.


```{r, eval=FALSE}
# Libs ----
if(!require(tidyverse)) {install.packages("tidyverse"); require(tidyverse)}
if(!require(sf)) {install.packages("sf"); require(sf)}
if(!require(arrow)) {install.packages("arrow"); require(arrow)}
if(!require(sfarrow)) {install.packages("sfarrow"); require(sfarrow)}
if(!require(terra)) {install.packages("terra"); require(terra)}
if(!require(raster)) {install.packages("raster"); require(raster)}
if(!require(fasterize)) {install.packages("fasterize"); require(fasterize)}
if(!require(gdalUtilities)){install.packages("gdalUtilities");require(gdalUtilities)}
if(!require(readxl)) {install.packages("readxl"); require(readxl)}

# templates ----
template_t=rast("./Templates/TemplateRasters/LV10m_10km.tif")
template_r=raster(template_t)


# class 200 ----

# topo
topo_udens_poly=st_read_parquet("./Geodata/2024/TopographicMap/HidroA_COMB.parquet")
topo_udens_poly=topo_udens_poly %>% 
  mutate(yes=200) %>% 
  dplyr::select(yes) %>% 
  st_transform(crs=3059)
topo_udens_lines=st_read_parquet("./Geodata/2024/TopographicMap/HidroL_COMB.parquet")
topo_udens_lines=topo_udens_lines %>% 
  mutate(yes=200) %>% 
  st_buffer(dist=5) %>% 
  dplyr::select(yes) %>% 
  st_transform(crs=3059)
topo_udens=rbind(topo_udens_poly,topo_udens_lines)
r_topo_udens=fasterize(topo_udens,template_r,field="yes")
raster::writeRaster(r_topo_udens,
                    "./RasterGrids_10m/2024/SimpleLandscape_class200_topo.tif",
                    progress="text")
# cleaning
rm(topo_udens_lines)
rm(topo_udens_poly)
rm(topo_udens)
rm(r_topo_udens)

# mkis
st_layers("./Geodata/2024/MKIS/MKIS_2025.gpkg")
mkis_gravji=st_read("./Geodata/2024/MKIS/MKIS_2025.gpkg",layer="Gravji")

mkis_gravji=mkis_gravji %>% 
  mutate(yes=200) %>% 
  st_buffer(dist=3) %>% 
  dplyr::select(yes)
r_mkis_udens=fasterize(mkis_gravji,template_r,field="yes")
raster::writeRaster(r_mkis_udens,
                    "./RasterGrids_10m/2024/SimpleLandscape_class200_mkis.tif",
                    progress="text")
# cleaning
rm(mkis_gravji)
rm(mkis_gravji2)
rm(mkis_gravji3)
rm(r_mkis_udens)

# lvm
lvm_gravji=st_read("./Geodata/2024/LVM_OpenData/LVM_GRAVJI/LVM_GRAVJI_Shape.shp")
lvm_gravji=lvm_gravji %>% 
  mutate(yes=200) %>% 
  st_buffer(dist=5) %>% 
  dplyr::select(yes)
r_lvm_gravji=fasterize(lvm_gravji,template_r,field="yes")
raster::writeRaster(r_lvm_gravji,
                    "./RasterGrids_10m/2024/SimpleLandscape_class200_lvm.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(lvm_gravji)
rm(r_lvm_gravji)


# merging
a200=rast("./RasterGrids_10m/2024/SimpleLandscape_class200_topo.tif")
b200=rast("./RasterGrids_10m/2024/SimpleLandscape_class200_mkis.tif")
c200=rast("./RasterGrids_10m/2024/SimpleLandscape_class200_lvm.tif")

udens_cover1=cover(a200,b200)
udens_cover2=cover(udens_cover1,c200,
                   filename="./RasterGrids_10m/2024/SimpleLandscape_class200_udens_premask.tif",
                   overwrite=TRUE)

# cleaning
rm(a200)
rm(b200)
rm(c200)
rm(udens_cover1)
rm(udens_cover2)
unlink("./RasterGrids_10m/2024/SimpleLandscape_class200_topo.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class200_mkis.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class200_lvm.tif")

```


- class `300` - **farmland**: agricultural land in LAD database **filled in 
sequence** - dominated by classes with higher values, but after the general 
classes were created, the gaps were filled in with information from Dynamic 
World. The following were combined to create this class:

– [LAD database](#Ch04.02), which, following the decision on grouping (classes are 
available [here](https://github.com/aavotins/HiQBioDiv_EGVs/blob/main/Data/Geodata/2024/LAD/KulturuKodi_2024.xlsx)), 
is divided into three broad groups (in order of overlap):

– arable land with class code `310`;

– fallow land with class code `320`;

– grassland with class code `330`;

– orchards and perennial shrub plantations in the general landscape are placed 
in other landscape classes.

The command lines below create a layer with landscape class `300` and its 
subclasses, which are saved in the file `SimpleLandscape_class300_lauki_premask.tif` 
for further processing.


```{r, eval=FALSE}
# Libs ----
if(!require(tidyverse)) {install.packages("tidyverse"); require(tidyverse)}
if(!require(sf)) {install.packages("sf"); require(sf)}
if(!require(arrow)) {install.packages("arrow"); require(arrow)}
if(!require(sfarrow)) {install.packages("sfarrow"); require(sfarrow)}
if(!require(terra)) {install.packages("terra"); require(terra)}
if(!require(raster)) {install.packages("raster"); require(raster)}
if(!require(fasterize)) {install.packages("fasterize"); require(fasterize)}
if(!require(gdalUtilities)){install.packages("gdalUtilities");require(gdalUtilities)}
if(!require(readxl)) {install.packages("readxl"); require(readxl)}

# templates ----
template_t=rast("./Templates/TemplateRasters/LV10m_10km.tif")
template_r=raster(template_t)


# class 300 ----

# lad
lad_klasem=read_excel("./Geodata/2024/LAD/KulturuKodi_2024.xlsx")
lad=st_read_parquet("./Geodata/2024/LAD/Lauki_2024.parquet")


## arable
amazemem=lad_klasem %>% 
  filter(str_detect(SDM_grupa_sakums,"aramz"))
aramzemes=lad %>% 
  filter(PRODUCT_CODE %in% amazemem$kods) %>% 
  mutate(yes=310) %>% 
  dplyr::select(yes)
r_aramzemes_lad=fasterize(aramzemes,template_r,field="yes")
raster::writeRaster(r_aramzemes_lad,
                    "./RasterGrids_10m/2024/SimpleLandscape_class310_aramzemes_lad.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(amazemem)
rm(aramzemes)
rm(r_aramzemes_lad)


## fallow
papuvem=lad_klasem %>% 
  filter(str_detect(SDM_grupa_sakums,"papuv"))
papuves=lad %>% 
  filter(PRODUCT_CODE %in% papuvem$kods) %>% 
  mutate(yes=320) %>% 
  dplyr::select(yes)
r_papuves_lad=fasterize(papuves,template_r,field="yes")
raster::writeRaster(r_papuves_lad,
                    "./RasterGrids_10m/2024/SimpleLandscape_class320_papuves_lad.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(papuvem)
rm(papuves)
rm(r_papuves_lad)

## grassland
zalajiem=lad_klasem %>% 
  filter(str_detect(SDM_grupa_sakums,"zālā"))
zalaji=lad %>% 
  filter(PRODUCT_CODE %in% zalajiem$kods) %>% 
  mutate(yes=330) %>% 
  dplyr::select(yes)
r_zalaji_lad=fasterize(zalaji,template_r,field="yes")
raster::writeRaster(r_zalaji_lad,
                    "./RasterGrids_10m/2024/SimpleLandscape_class330_zalaji_lad.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(zalajiem)
rm(zalaji)
rm(r_zalaji_lad)

# merging
a300=rast("./RasterGrids_10m/2024/SimpleLandscape_class310_aramzemes_lad.tif")
b300=rast("./RasterGrids_10m/2024/SimpleLandscape_class320_papuves_lad.tif")
c300=rast("./RasterGrids_10m/2024/SimpleLandscape_class330_zalaji_lad.tif")

farmland_cover1=cover(a300,b300)
farmland_cover2=cover(farmland_cover1,c300,
                          filename="./RasterGrids_10m/2024/SimpleLandscape_class300_lauki_premask.tif",
                          overwrite=TRUE)
# cleaning
rm(lad)
rm(lad_klasem)
rm(a300)
rm(b300)
rm(c300)
rm(farmland_cover1)
rm(farmland_cover2)
unlink("./RasterGrids_10m/2024/SimpleLandscape_class310_aramzemes_lad.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class320_papuves_lad.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class330_zalaji_lad.tif")

```


- class `400` - **allotment gardens and orchards, cottages**, **filled in order** - 
dominate classes with higher values. To create this class, the following were 
combined (in order of overlap):

– [topographic map](#Ch04.04) layer `LandusA_COMB`, the result of which is coded with `410`;

– [LAD database](#Ch04.02) rural information layer group (classes are 
available [here](https://github.com/aavotins/HiQBioDiv_EGVs/blob/main/Data/Geodata/2024/LAD/KulturuKodi_2024.xlsx)) 
“orchards”, the result of which is coded with `420`.

The command lines below create a layer with landscape class `400`, which is saved 
in the file `SimpleLandscape_class400_vasarnicas_premask.tif` for further 
processing.


```{r, eval=FALSE}
# Libs ----
if(!require(tidyverse)) {install.packages("tidyverse"); require(tidyverse)}
if(!require(sf)) {install.packages("sf"); require(sf)}
if(!require(arrow)) {install.packages("arrow"); require(arrow)}
if(!require(sfarrow)) {install.packages("sfarrow"); require(sfarrow)}
if(!require(terra)) {install.packages("terra"); require(terra)}
if(!require(raster)) {install.packages("raster"); require(raster)}
if(!require(fasterize)) {install.packages("fasterize"); require(fasterize)}
if(!require(gdalUtilities)){install.packages("gdalUtilities");require(gdalUtilities)}
if(!require(readxl)) {install.packages("readxl"); require(readxl)}

# templates ----
template_t=rast("./Templates/TemplateRasters/LV10m_10km.tif")
template_r=raster(template_t)


# class 400 ----

# topo
darzini_topo=st_read_parquet("./Geodata/2024/TopographicMap/LandusA_COMB.parquet")
table(darzini_topo$FNAME,useNA="always")
darzini_topo=darzini_topo %>% 
  filter(FNAME %in% c("poligons_Augludarzs","poligons_Augļudārzs","poligons_Sakņudārzs",
                      "poligons_Ogulājs","poligons_Ogulajs","poligons_Saknudarzs")) %>% 
  mutate(yes=410) %>% 
  dplyr::select(yes)
r_darzini_topo=fasterize(darzini_topo,template_r,field="yes")
raster::writeRaster(r_darzini_topo,
                    "./RasterGrids_10m/2024/SimpleLandscape_class410_darzini_topo.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(darzini_topo)
rm(r_darzini_topo)

# lad
lad_klasem=read_excel("./Geodata/2024/LAD/KulturuKodi_2024.xlsx")
table(lad_klasem$SDM_grupa_sakums,useNA="always")
augludarziem=lad_klasem %>% 
  filter(SDM_grupa_sakums=="augļudārzi")
lad=st_read_parquet("./Geodata/2024/LAD/Lauki_2024.parquet")
lad=lad %>% 
  filter(PRODUCT_CODE %in% augludarziem$kods) %>% 
  mutate(yes=420) %>% 
  dplyr::select(yes)
r_darzini_lad=fasterize(lad,template_r,field="yes")
raster::writeRaster(r_darzini_lad,
                    "./RasterGrids_10m/2024/SimpleLandscape_class420_darzini_lad.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(lad_klasem)
rm(augludarziem)
rm(lad)
rm(r_darzini_lad)

# merging
a400=rast("./RasterGrids_10m/2024/SimpleLandscape_class410_darzini_topo.tif")
b400=rast("./RasterGrids_10m/2024/SimpleLandscape_class420_darzini_lad.tif")

allotment_cover=cover(a400,b400,
                     filename="./RasterGrids_10m/2024/SimpleLandscape_class400_varnicas_premask.tif",
                     overwrite=TRUE)

# cleaning
rm(a400)
rm(b400)
rm(allotment_cover)
unlink("./RasterGrids_10m/2024/SimpleLandscape_class410_darzini_topo.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class420_darzini_lad.tif")

```

- class `500` - **built-up**: built-up areas, filled in at the end (see section 
"merging and filling" of this chapter) using information from Dynamic World about 
places not covered by other classes.

- class `600` - **forests, shrublands, clearings**: areas covered with trees and 
shrubs, clearings, and dead forest stands, **filled in order** - dominates 
classes with higher values. The following have been combined to create this 
class (in order of overlap):

– [The Global Forest Watch](#Ch04.09) layer records of tree canopy cover 
loss since 2020, coded as `610`;

– [Forest State Register](#Ch04.01) clearings and dead forest stands, the result 
of which is coded as `610`;

– [Forest State Register](#Ch04.01) marked forest stands that are lower than 5 m 
and seed production plantations, the result of which is coded as `620`;

– [topographic map](#Ch04.04) layer `FloraL_COMB` classes related to shrubs, 
buffered by 10 m, coded as `620`;

– [topographic map](#Ch04.04) layers `LandusA_COMB` clases "poligons_Krūmājs", 
"poligons_Krumajs", "poligons_Krūmaugu_plant", "poligons_Plantacija_krum", coded as `620`;

– [LAD database](#Ch04.02) group (classes are 
available [here](https://github.com/aavotins/HiQBioDiv_EGVs/blob/main/Data/Geodata/2024/LAD/KulturuKodi_2024.xlsx)) 
“krūmveida ilggadīgie stādījumi”, the result of which is coded with `620`;

– [Forest State Register](#Ch04.01) forest stands with a height of at least 5 m, 
coded as `630`;

– [topographic map](#Ch04.04) layer `LandusA_COMB` classes "poligons_Parks", 
"poligons_Meza_kapi", "poligons_Kapi", "poligons_Kapi_meza", the result of 
which is coded as `640`;

– [topographic map](#Ch04.04) layer `FloraL_COMB` with tree-related classes 
buffered by 10 m, coded as `640`;

– [Palsar Forests](#Ch04.10) layer, coded as `630`.

The command lines below create a layer with landscape class `600`, which is saved 
in the file `SimpleLandscape_class600_meziem_premask.tif` for further processing.


```{r, eval=FALSE}
# Libs ----
if(!require(tidyverse)) {install.packages("tidyverse"); require(tidyverse)}
if(!require(sf)) {install.packages("sf"); require(sf)}
if(!require(arrow)) {install.packages("arrow"); require(arrow)}
if(!require(sfarrow)) {install.packages("sfarrow"); require(sfarrow)}
if(!require(terra)) {install.packages("terra"); require(terra)}
if(!require(raster)) {install.packages("raster"); require(raster)}
if(!require(fasterize)) {install.packages("fasterize"); require(fasterize)}
if(!require(gdalUtilities)){install.packages("gdalUtilities");require(gdalUtilities)}
if(!require(readxl)) {install.packages("readxl"); require(readxl)}

# templates ----
template_t=rast("./Templates/TemplateRasters/LV10m_10km.tif")
template_r=raster(template_t)


# class 600 ----

# mvr 
mvr=st_read_parquet("./Geodata/2024/MVR/nogabali_2024janv.parquet")

# clearcuts
izcirtumi=mvr %>% 
  filter(zkat %in% c("12","14")) %>% 
  mutate(yes=610) %>% 
  dplyr::select(yes)
r_izcirtumi_mvr=fasterize(izcirtumi,template_r,field="yes")
raster::writeRaster(r_izcirtumi_mvr,
                    "./RasterGrids_10m/2024/SimpleLandscape_class610_izcirtumi_mvr.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(izcirtumi)
rm(r_izcirtumi_mvr)

# low stands
# also zkat 16
zemas_audzes=mvr %>% 
  filter((zkat =="10" & h10<5)|zkat=="16") %>% 
  mutate(yes=620) %>% 
  dplyr::select(yes)
r_zemas_mvr=fasterize(zemas_audzes,template_r,field="yes")
raster::writeRaster(r_zemas_mvr,
                    "./RasterGrids_10m/2024/SimpleLandscape_class620_zemas_mvr.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(zemas_audzes)
rm(r_zemas_mvr)


# high stands
augstas_audzes=mvr %>% 
  filter(zkat =="10" & h10>=5) %>% 
  mutate(yes=630) %>% 
  dplyr::select(yes)
r_augstas_mvr=fasterize(augstas_audzes,template_r,field="yes")
raster::writeRaster(r_augstas_mvr,
                    "./RasterGrids_10m/2024/SimpleLandscape_class630_augstas_mvr.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(augstas_audzes)
rm(r_augstas_mvr)
rm(mvr)

# tcl - since 2020
tcl=rast("./Geodata/2024/Trees/GFW/TreeCoverLoss_v1_12.tif")
tcl2=ifel(tcl<20,NA,610,
          filename="./RasterGrids_10m/2024/SimpleLandscape_class610_TCL.tif",
          overwrite=TRUE)
# cleaning
rm(tcl)
rm(tcl2)

# palsar
palsar=rast("./Geodata/2024/Trees/Palsar/Palsar_Forests.tif")
palsar2=ifel(palsar==1,630,NA,
             filename="./RasterGrids_10m/2024/SimpleLandscape_class630_Palsar.tif",
             overwrite=TRUE)
# cleaning
rm(palsar)
rm(palsar2)


# lad
lad_klasem=read_excel("./Geodata/2024/LAD/KulturuKodi_2024.xlsx")
table(lad_klasem$SDM_grupa_sakums,useNA="always")
lad=st_read_parquet("./Geodata/2024/LAD/Lauki_2024.parquet")
krumiem=lad_klasem %>% 
  filter(str_detect(SDM_grupa_sakums,"krūmv"))
krumi=lad %>% 
  filter(PRODUCT_CODE %in% krumiem$kods) %>% 
  mutate(yes=620) %>% 
  dplyr::select(yes)
r_krumi_lad=fasterize(krumi,template_r,field="yes")
raster::writeRaster(r_krumi_lad,
                    "./RasterGrids_10m/2024/SimpleLandscape_class620_krumi_lad.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(lad_klasem)
rm(lad)
rm(krumiem)
rm(krumi)
rm(r_krumi_lad)

# topo - pkk
pkk_topo=st_read_parquet("./Geodata/2024/TopographicMap/LandusA_COMB.parquet")
table(pkk_topo$FNAME,useNA="always")
pkk_topo=pkk_topo %>% 
  filter(FNAME %in% c("poligons_Parks","poligons_Meza_kapi","poligons_Kapi",
                      "poligons_Kapi_meza")) %>% 
  mutate(yes=640) %>% 
  dplyr::select(yes)
r_pkk_topo=fasterize(pkk_topo,template_r,field="yes")
raster::writeRaster(r_pkk_topo,
                    "./RasterGrids_10m/2024/SimpleLandscape_class640_pkk_topo.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(pkk_topo)
rm(r_pkk_topo)

# topo - shrubs
krumi_topo=st_read_parquet("./Geodata/2024/TopographicMap/LandusA_COMB.parquet")
table(krumi_topo$FNAME,useNA="always")
krumi_topo=krumi_topo %>% 
  filter(FNAME %in% c("poligons_Krūmājs","poligons_Krumajs",
                      "poligons_Krūmaugu_plant","poligons_Plantacija_krum")) %>% 
  mutate(yes=620) %>% 
  dplyr::select(yes)
r_krumi_topo=fasterize(krumi_topo,template_r,field="yes")
raster::writeRaster(r_krumi_topo,
                    "./RasterGrids_10m/2024/SimpleLandscape_class620_krumi_topo.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(krumi_topo)
rm(r_krumi_topo)

# topo - linear vegetation
linijas_topo=st_read_parquet("./Geodata/2024/TopographicMap/FloraL_COMB.parquet")
table(linijas_topo$FNAME,useNA="always")

# linear shrubs
krumu_linijas_topo=linijas_topo %>% 
  filter(FNAME=="Krūmu rinda dzīvzogs"|FNAME=="Krūmu rinda gar ceļiem upēm"|
           FNAME=="Krumu_rinda_dzivzogs"|FNAME=="Krumu_rinda_gar_celiem_upem") %>% 
  mutate(yes=620) %>% 
  st_buffer(dist=10) %>% 
  dplyr::select(yes)
r_krumu_linijas_topo=fasterize(krumu_linijas_topo,template_r,field="yes")
raster::writeRaster(r_krumu_linijas_topo,
                    "./RasterGrids_10m/2024/SimpleLandscape_class620_KrumuLinijas_topo.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(krumu_linijas_topo)
rm(r_krumu_linijas_topo)

# linear trees
koku_linijas_topo=linijas_topo %>% 
  filter(str_detect(FNAME,"Koku")) %>% 
  mutate(yes=640) %>% 
  st_buffer(dist=10) %>% 
  dplyr::select(yes)
r_koku_linijas_topo=fasterize(koku_linijas_topo,template_r,field="yes")
raster::writeRaster(r_koku_linijas_topo,
                    "./RasterGrids_10m/2024/SimpleLandscape_class640_KokuLinijas_topo.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(koku_linijas_topo)
rm(r_koku_linijas_topo)
rm(linijas_topo)

# merging
r_krumi_lad=rast("./RasterGrids_10m/2024/SimpleLandscape_class620_krumi_lad.tif")
r_pkk_topo=rast("./RasterGrids_10m/2024/SimpleLandscape_class640_pkk_topo.tif")
r_krumi_topo=rast("./RasterGrids_10m/2024/SimpleLandscape_class620_krumi_topo.tif")
r_krumu_linijas_topo=rast("./RasterGrids_10m/2024/SimpleLandscape_class620_KrumuLinijas_topo.tif")
r_koku_linijas_topo=rast("./RasterGrids_10m/2024/SimpleLandscape_class640_KokuLinijas_topo.tif")
r_palsar=rast("./RasterGrids_10m/2024/SimpleLandscape_class630_palsar.tif")
r_tcl=rast("./RasterGrids_10m/2024/SimpleLandscape_class610_TCL.tif")
r_augstas_mvr=rast("./RasterGrids_10m/2024/SimpleLandscape_class630_augstas_mvr.tif")
r_zemas_mvr=rast("./RasterGrids_10m/2024/SimpleLandscape_class620_zemas_mvr.tif")
r_izcirtumi_mvr=rast("./RasterGrids_10m/2024/SimpleLandscape_class610_izcirtumi_mvr.tif")


mezu_cover=cover(r_tcl,r_izcirtumi_mvr)
mezu_cover=cover(mezu_cover,r_zemas_mvr)
mezu_cover=cover(mezu_cover,r_krumu_linijas_topo)
mezu_cover=cover(mezu_cover,r_krumi_topo)
mezu_cover=cover(mezu_cover,r_krumi_lad)
mezu_cover=cover(mezu_cover,r_augstas_mvr)
mezu_cover=cover(mezu_cover,r_pkk_topo)
mezu_cover=cover(mezu_cover,r_koku_linijas_topo)
mezu_cover=cover(mezu_cover,r_palsar,
                 filename="./RasterGrids_10m/2024/SimpleLandscape_class600_meziem_premask.tif",
                 overwrite=TRUE)


# cleaning
rm(r_krumi_lad)
rm(r_pkk_topo)
rm(r_krumi_topo)
rm(r_krumu_linijas_topo)
rm(r_koku_linijas_topo)
rm(r_palsar)
rm(r_tcl)
rm(r_augstas_mvr)
rm(r_zemas_mvr)
rm(r_izcirtumi_mvr)
rm(mezu_cover)

unlink("./RasterGrids_10m/2024/SimpleLandscape_class620_krumi_lad.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class640_pkk_topo.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class620_krumi_topo.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class620_KrumuLinijas_topo.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class640_KokuLinijas_topo.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class630_palsar.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class610_TCL.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class630_augstas_mvr.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class620_zemas_mvr.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class610_izcirtumi_mvr.tif")

```

- class `700` - **wetlands**: combining geospatial data related to reed beds, 
marshes, mires and bogs, **filled in order except class `720` that dominates over waters** - 
dominates classes with higher values. To create this class, the following were 
combined (in order of overlap):

– [topographic map](#Ch04.04) layer `LandusA_COMB` classes "Meldrājs_ūdenī_poligons", 
"poligons_Grislajs", "poligons_Grīslājs", "poligons_Meldrajs", "poligons_Meldrājs", 
"poligons_Meldrajs_udeni", "poligons_Nec_purvs_grīslājs", "poligons_Nec_purvs_meldrājs", 
"Sēklis_poligons", the result of which is coded with `720`;

– [topographic map](#Ch04.04) layer `LandusA_COMB` class "poligons_Nec_purvs_sūnājs", 
"poligons_Sunajs", "poligons_Sūnājs", the result of which is coded with `710`;

– [topographic map](#Ch04.04) layer `SwampA_COMB`, the result of which is coded 
as `710`;

– land categories “21”, “22”, “23” marked in the [State Forest Register](#Ch04.01), 
the result of which is coded as `710`;

– land categories “41” and “42” marked in the [State Forest Register](#Ch04.01), 
the result of which is coded as `730`;

- bogs from [Bogs and Mires: EDI](#Ch04.17);

- transitional mires from [Bogs and Mires: EDI](#Ch04.17);

The command lines below create a layer with landscape class `700`, which is saved 
in the file `SimpleLandscape_class700_mitraji_premask.tif` for further processing.


```{r, eval=FALSE}
# Libs ----
if(!require(tidyverse)) {install.packages("tidyverse"); require(tidyverse)}
if(!require(sf)) {install.packages("sf"); require(sf)}
if(!require(arrow)) {install.packages("arrow"); require(arrow)}
if(!require(sfarrow)) {install.packages("sfarrow"); require(sfarrow)}
if(!require(terra)) {install.packages("terra"); require(terra)}
if(!require(raster)) {install.packages("raster"); require(raster)}
if(!require(fasterize)) {install.packages("fasterize"); require(fasterize)}
if(!require(gdalUtilities)){install.packages("gdalUtilities");require(gdalUtilities)}
if(!require(readxl)) {install.packages("readxl"); require(readxl)}

# templates ----
template_t=rast("./Templates/TemplateRasters/LV10m_10km.tif")
template_r=raster(template_t)


# class 700 ----

# topo
topo=st_read_parquet("./Geodata/2024/TopographicMap/LandusA_COMB.parquet")
table(topo$FNAME,useNA="always")

## ReedSedgeRush
niedraji_topo=topo %>% 
  filter(FNAME %in% c("Meldrājs_ūdenī_poligons","poligons_Grislajs","poligons_Grīslājs",
                      "poligons_Meldrajs","poligons_Meldrājs","poligons_Meldrajs_udeni",
                      "poligons_Nec_purvs_grīslājs",
                      "poligons_Nec_purvs_meldrājs",
                      "Sēklis_poligons")) %>% 
  mutate(yes=720) %>% 
  dplyr::select(yes)
r_niedraji_topo=fasterize(niedraji_topo,template_r,field="yes")
raster::writeRaster(r_niedraji_topo,
                    "./RasterGrids_10m/2024/SimpleLandscape_class720_niedraji_topo.tif",
                    progress="text")
# cleaning
rm(niedraji_topo)
rm(r_niedraji_topo)


## bogs
purvi_topo=topo %>% 
  filter(FNAME %in% c("poligons_Nec_purvs_sūnājs",
                      "poligons_Sunajs","poligons_Sūnājs")) %>% 
  mutate(yes=710) %>% 
  dplyr::select(yes)
topo_purvi=st_read_parquet("./Geodata/2024/TopographicMap/SwampA_COMB.parquet")
topo_purvi=topo_purvi %>% 
  mutate(yes=710) %>% 
  dplyr::select(yes)
purvi=rbind(purvi_topo,topo_purvi)
r_purvi_topo=fasterize(purvi,template_r,field="yes")
raster::writeRaster(r_purvi_topo,
                    "./RasterGrids_10m/2024/SimpleLandscape_class710_purvi_topo.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(purvi_topo)
rm(topo_purvi)
rm(purvi)
rm(r_purvi_topo)


# mvr
mvr=st_read_parquet("./Geodata/2024/MVR/nogabali_2024janv.parquet")

# bogs and mires
mvr_purvi=mvr %>% 
  filter(zkat %in% c("21","22","23")) %>% 
  mutate(yes=710) %>% 
  dplyr::select(yes)
r_purvi_mvr=fasterize(mvr_purvi,template_r,field="yes")
raster::writeRaster(r_purvi_mvr,
                    "./RasterGrids_10m/2024/SimpleLandscape_class710_purvi_mvr.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(mvr_purvi)
rm(r_purvi_mvr)

# beavers
mvr_bebri=mvr %>% 
  filter(zkat %in% c("41","42")) %>% 
  mutate(yes=730) %>% 
  dplyr::select(yes)
r_bebri_mvr=fasterize(mvr_bebri,template_r,field="yes")
raster::writeRaster(r_bebri_mvr,
                    "./RasterGrids_10m/2024/SimpleLandscape_class730_bebri_mvr.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(mvr_bebri)
rm(r_bebri_mvr)
rm(mvr)



# merging
r_niedraji_topo=rast("./RasterGrids_10m/2024/SimpleLandscape_class720_niedraji_topo.tif")
r_purvi_topo=rast("./RasterGrids_10m/2024/SimpleLandscape_class710_purvi_topo.tif")
r_purvi_mvr=rast("./RasterGrids_10m/2024/SimpleLandscape_class710_purvi_mvr.tif")
r_bebri_mvr=rast("./RasterGrids_10m/2024/SimpleLandscape_class730_bebri_mvr.tif")
mires=rast("./RasterGrids_10m/2024/EDI_TransitionalMiresYN.tif")
bogs=rast("./RasterGrids_10m/2024/EDI_BogsYN.tif")

wetlands_cover=cover(r_niedraji_topo,r_purvi_topo)
wetlands_cover=cover(wetlands_cover,r_purvi_mvr)
wetlands_cover=cover(wetlands_cover,r_bebri_mvr)
wetlands_cover=cover(wetlands_cover,mires)
wetlands_cover=cover(wetlands_cover,bogs,
                            filename="./RasterGrids_10m/2024/SimpleLandscape_class700_mitraji_premask.tif",
                            overwrite=TRUE)
# cleaning
rm(r_niedraji_topo)
rm(r_purvi_topo)
rm(r_purvi_mvr)
rm(r_bebri_mvr)
rm(bogs)
rm(mires)
rm(topo)
rm(wetlands_cover)

unlink("./RasterGrids_10m/2024/SimpleLandscape_class710_purvi_topo.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class710_purvi_mvr.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class730_bebri_mvr.tif")

```

- class `800` - **bare soil and quarries**: combining layers related to bare soil, 
heaths, and quarries, **filled in order** - as this is the highest class, it 
dominates only over Dynamic World used to fill gaps. The following have been 
combined to create this class (in order of overlap):

– [topographic map](#Ch04.04) layer `LandusA_COMB` classes 
"poligons_Smiltājs", "poligons_Smiltajs", "poligons_Grants", "poligons_Kūdra", "poligons_Virsajs" 
the result of which is coded 
with `800`;

– land categories "33" and "34" marked in the [State Forest Register](#Ch04.01), 
the result of which is coded as `730`.

The command lines below create a layer with landscape class `800`, which is saved 
in the file `SimpleLandscape_class800_smiltaji_premask.tif` for further processing.


```{r, eval=FALSE}
# Libs ----
if(!require(tidyverse)) {install.packages("tidyverse"); require(tidyverse)}
if(!require(sf)) {install.packages("sf"); require(sf)}
if(!require(arrow)) {install.packages("arrow"); require(arrow)}
if(!require(sfarrow)) {install.packages("sfarrow"); require(sfarrow)}
if(!require(terra)) {install.packages("terra"); require(terra)}
if(!require(raster)) {install.packages("raster"); require(raster)}
if(!require(fasterize)) {install.packages("fasterize"); require(fasterize)}
if(!require(gdalUtilities)){install.packages("gdalUtilities");require(gdalUtilities)}
if(!require(readxl)) {install.packages("readxl"); require(readxl)}

# templates ----
template_t=rast("./Templates/TemplateRasters/LV10m_10km.tif")
template_r=raster(template_t)


# class 800 ----

smiltaji_topo=st_read_parquet("./Geodata/2024/TopographicMap/LandusA_COMB.parquet")
table(smiltaji_topo$FNAME,useNA="always")
smiltaji_topo=smiltaji_topo %>% 
  filter(FNAME %in% c("poligons_Smiltājs","poligons_Smiltajs","poligons_Grants",
                      "poligons_Kūdra","poligons_Virsajs")) %>% 
  mutate(yes=800) %>% 
  dplyr::select(yes)
r_smiltaji_topo=fasterize(smiltaji_topo,template_r,field="yes")
raster::writeRaster(r_smiltaji_topo,
                    "./RasterGrids_10m/2024/SimpleLandscape_class800_SmiltajiKudra_topo.tif",
                    progress="text")
# cleaning
rm(smiltaji_topo)
rm(r_smiltaji_topo)

# mvr zkat 33 un 34
mvr=st_read_parquet("./Geodata/2024/MVR/nogabali_2024janv.parquet")

smiltajiem=mvr %>% 
  filter(zkat %in% c("33","34")) %>% 
  mutate(yes=800) %>% 
  dplyr::select(yes)
r_smiltaji_mvr=fasterize(smiltajiem,template_r,field="yes")
raster::writeRaster(r_smiltaji_mvr,
                    "./RasterGrids_10m/2024/SimpleLandscape_class800_SmiltVirs_mvr.tif",
                    progress="text",
                    overwrite=TRUE)
# cleaning
rm(mvr)
rm(smiltajiem)
rm(r_smiltaji_mvr)

# merging
r_smiltaji_topo=rast("./RasterGrids_10m/2024/SimpleLandscape_class800_SmiltajiKudra_topo.tif")
r_smiltaji_mvr=rast("./RasterGrids_10m/2024/SimpleLandscape_class800_SmiltVirs_mvr.tif")

bare_cover=terra::merge(r_smiltaji_topo,r_smiltaji_mvr,
                               filename="./RasterGrids_10m/2024/SimpleLandscape_class800_smiltaji_premask.tif",
                               overwrite=TRUE)
# liekā aizvākšana
rm(r_smiltaji_topo)
rm(r_smiltaji_mvr)
rm(bare_cover)

unlink("./RasterGrids_10m/2024/SimpleLandscape_class800_SmiltajiKudra_topo.tif")
unlink("./RasterGrids_10m/2024/SimpleLandscape_class800_SmiltVirs_mvr.tif")

```

**Merging and filling**

The command lines below combine the previously created layers with landscape 
classes in the correct order and ensure that gaps are filled with the appropriately 
classified Dynamic World composite for April-August 2024. After masking to only 
the analysis space, layer is saved in the in the file `Ainava_vienk_mask.tif` 
for further processing.

```{r, eval=FALSE}
# Libs ----
if(!require(tidyverse)) {install.packages("tidyverse"); require(tidyverse)}
if(!require(sf)) {install.packages("sf"); require(sf)}
if(!require(arrow)) {install.packages("arrow"); require(arrow)}
if(!require(sfarrow)) {install.packages("sfarrow"); require(sfarrow)}
if(!require(terra)) {install.packages("terra"); require(terra)}
if(!require(raster)) {install.packages("raster"); require(raster)}
if(!require(fasterize)) {install.packages("fasterize"); require(fasterize)}
if(!require(gdalUtilities)){install.packages("gdalUtilities");require(gdalUtilities)}
if(!require(readxl)) {install.packages("readxl"); require(readxl)}

# templates ----
template_t=rast("./Templates/TemplateRasters/LV10m_10km.tif")
template_r=raster(template_t)



# final merging and covering ----

# DW  
dynworld=rast("Geodata/2024/DynamicWorld/DW_2024_apraug.tif")
klases=matrix(c(0,200,
                1,620,
                2,330,
                3,720,
                4,310,
                5,710,
                6,500,
                7,800,
                8,500),ncol=2,byrow=TRUE)
dw2=terra::classify(dynworld,klases)
writeRaster(dw2,
            "./RasterGrids_10m/2024/DW_reclass.tif",
            overwrite=TRUE)
# other layers
celi=rast("./RasterGrids_10m/2024/SimpleLandscape_class100_celi.tif")
niedraji=rast("RasterGrids_10m/2024/SimpleLandscape_class720_niedraji_topo.tif")
udeni=rast("./RasterGrids_10m/2024/SimpleLandscape_class200_udens_premask.tif")
lauki=rast("./RasterGrids_10m/2024/SimpleLandscape_class300_lauki_premask.tif")
vasarnicas=rast("./RasterGrids_10m/2024/SimpleLandscape_class400_varnicas_premask.tif")
mezi=rast("./RasterGrids_10m/2024/SimpleLandscape_class600_meziem_premask.tif")
mitraji=rast("./RasterGrids_10m/2024/SimpleLandscape_class700_mitraji_premask.tif")
smiltaji=rast("./RasterGrids_10m/2024/SimpleLandscape_class800_smiltaji_premask.tif")
dw2=rast("./RasterGrids_10m/2024/DW_reclass.tif")

# covering in correct order
rastri_ainavai=cover(celi,niedraji)
rastri_ainavai=cover(rastri_ainavai,udeni)
rastri_ainavai=cover(rastri_ainavai,lauki)
rastri_ainavai=cover(rastri_ainavai,vasarnicas)
rastri_ainavai=cover(rastri_ainavai,mezi)
rastri_ainavai=cover(rastri_ainavai,mitraji)
rastri_ainavai=cover(rastri_ainavai,smiltaji)
rastri_ainavai=cover(rastri_ainavai,dw2,
                           filename="./RasterGrids_10m/2024/Ainava_vienkarsa.tif",
                           overwrite=TRUE)

# cleaning
rm(celi)
rm(niedraji)
rm(udeni)
rm(lauki)
rm(vasarnicas)
rm(mezi)
rm(mitraji)
rm(smiltaji)
rm(klases)
rm(dynworld)
rm(dw2)
rm(rastri_ainavai)

# masking
rastrs_ainava=rast("./RasterGrids_10m/2024/Ainava_vienkarsa.tif")
masketa_ainava=terra::mask(rastrs_ainava,
                           template_t,
                           filename="./RasterGrids_10m/2024/Ainava_vienk_mask.tif",
                           overwrite=TRUE)
plot(masketa_ainava)

# cleaning
rm(rastrs_ainava)
rm(masketa_ainava)

```





## Landscape diversity {#Ch05.04}

krā

### Landscape in general diversity {#Ch05.04.01}

krā


### Forest diversity {#Ch05.04.02}

krā

### Farmland diversity {#Ch05.04.03}

krā


